---
title: "Prediction using Posterior"
format: html
---

**Objectives**:

-   Understand and implement the concept of posterior predictive power in Bayesian Statistics using Markov Chain values.

-   Learn to use R and relevant libraries (e.g., *Stan*, *bayesian* and *ggplot2*) to approximate posterior predictive power.

-   Apply the concept to real-world examples from the medical field.

**Materials Needed:**

-   RStudio or any R environment

-   R packages: *rstan*, *bayesian*, *ggplot2*

-   Example medical datasets (e.g., clinical trial results, patient outcomes)

**Introduction:**

-   Recap Bayesian Statistics basics, emphasizing posterior distribution.

-   Introduce the concept of posterior predictive power and its importance in model assessment.

-   Briefly explain Markov Chain values and their role in Bayesian modeling.

**Markov Chain Monte Carlo (MCMC) and Bayesian Modeling:**

-   Provide an overview of Markov Chain Monte Carlo methods.

-   Discuss Bayesian modeling using *R*, highlighting the *rstan* package.

-   Demonstrate the process of obtaining samples from the posterior distribution.

**Posterior Predictive Power:**

-   Define posterior predictive power and distinguish between prior and posterior predictive checks.

-   Emphasize the need for assessing the model's ability to generate new data.

-   Discuss the limitations of relying solely on parameter estimates.

**Using Markov Chain Values:**

-   Walk through the steps of using Markov Chain values to approximate posterior predictive power.

-   Provide R code snippets for extracting Markov Chain values using *rstan*.

-   Illustrate the importance of visualizing the posterior predictive distribution.

```{r}

# Example R code for using Markov Chain values to approximate posterior predictive power

# Install and load necessary packages
# install.packages(c("rstan", "bayesian", "ggplot2"))
# library(rstan)
# library(bayesian)
# library(ggplot2)

# Load your medical dataset (replace "your_data.csv" with the actual file name)
# data <- read.csv("your_data.csv")

# Define the Bayesian model using Stan
# stan_code <- "
# data {
# int<lower=0> N;
  # Include your data variables here
  # Example: real y[N];
# }

# parameters {
  # Include your model parameters here
  # Example: real mu;
# }

# model {
  # Include your prior and likelihood specifications here
  # Example: y ~ normal(mu, sigma);
# }

# generated quantities {
 # Simulate new data points using posterior samples
 # Example: real y_pred[N];
 # for (i in 1:N) {
 # y_pred[i] = normal_rng(mu, sigma);
 # }
# }


# Compile the Stan model
# stan_model <- stan_model(model_code = stan_code)

# Fit the model to your data
# stan_fit <- sampling(stan_model, data = list(N = nrow(data), y = data$your_variable))

# Extract Markov Chain values
# chain_values <- as.matrix(stan_fit)

# Visualize posterior predictive distribution
# posterior_predictive <- bayesian::extract_posterior_predictive(stan_fit)
# ggplot(data, aes(x = your_variable)) +
#  geom_density() +
#  geom_density(data = posterior_predictive, aes(y = ..scaled..), color = "blue", alpha = 0.5) +
# labs(title = "Posterior Predictive Distribution", x = "Your Variable")

```

**Example from the Medical Field:**

-   Introduce a medical dataset and guide participants through the process of building a Bayesian model.

-   Discuss model assessment and emphasize the use of Markov Chain values.

-   Provide hands-on exercises for participants to implement the concepts using their own medical datasets.

**Hands-On Exercise:**

-   Students will work on a provided medical dataset or bring their own.

-   Instructor will guide students through building a Bayesian model, obtaining Markov Chain values, and assessing posterior predictive power.

-   Encourage discussion and troubleshooting.

**Conclusion:**

-   Recap key points about Bayesian modeling, posterior predictive power, and the use of Markov Chain values in R.

-   Discuss the relevance and applications of these techniques in the medical field.

-   Encourage participants to explore additional resources for further learning.

**Assessment:**

-   Evaluate students’ understanding through class participation, engagement in the hands-on exercise, and the quality of their interpretations of posterior predictive power in the medical context.

-------------------------------------------------------------------------------------------------------------------

**Additional Notes**

**Posterior Predictive Power**

Posterior predictive power is a concept in Bayesian statistics that involves assessing the ability of a statistical model to generate new data that is consistent with observed data. It is a crucial aspect of model assessment in Bayesian statistics, as it goes beyond simply fitting a model to observed data and focuses on the model's ability to make predictions for future or unseen data.

Here's a breakdown of the key components:

***Posterior Predictive Distribution***:

In Bayesian statistics, the posterior predictive distribution is obtained by combining the likelihood function (which describes how well the model explains the observed data) with the posterior distribution of the model parameters (updated beliefs about parameters based on both prior information and the observed data). Mathematically, it is expressed as:

-   P(future data∣observed data)

=∫P(future data∣parameters)⋅(parameters∣observed data) P(parameters

-   P(future data∣observed data)=∫P(future data∣parameters)⋅P(parameters∣observed data) d parameters

-   This distribution represents the model's prediction for future data given the observed data.

    ***Posterior Predictive Power***:

-   Posterior predictive power is a measure of how well the model's predictions align with new, unseen data. It helps evaluate the model's ability to generalize beyond the observed data.

```{=html}
<!-- -->
```
-   It can be assessed using various measures, such as predictive accuracy, calibration, and other model performance metrics.

    ***Importance in Model Assessment***:

    -   Model fitting alone does not guarantee that a model will perform well on new data. Posterior predictive power provides a more comprehensive evaluation by considering both the fit to observed data and the model's predictive capabilities.

    -   It helps identify whether the model is overfitting (fitting noise in the data rather than capturing underlying patterns) or underfitting (oversimplifying the data).

    **Examples from the Medical Field**:

    -   ***Clinical Trials***: In medical research, Bayesian models are often used to analyze clinical trial data. Posterior predictive power can be employed to assess how well the model predicts patient outcomes or responses to treatment in new trials.

    -   ***Diagnostic Models***: For medical diagnostic models, posterior predictive power can be used to evaluate how well the model predicts outcomes for patients not included in the initial dataset. This is crucial for assessing the model's generalizability in real-world scenarios.

    -   ***Epidemiological Models***: Bayesian models are used in epidemiology to understand disease spread and predict future outbreaks. Posterior predictive power helps assess the reliability of these models in forecasting the dynamics of infectious diseases.

    -   ***Health Outcome Predictions***: In personalized medicine, Bayesian models can be used to predict individual health outcomes based on genetic, clinical, and environmental factors. Posterior predictive power is essential to evaluate the accuracy of these predictions.

In summary, posterior predictive power in Bayesian statistics provides a robust framework for assessing the overall performance of models, especially in fields like medicine where accurate predictions are crucial for patient well-being and decision-making.

#### **Markov Chain Monte Carlo (MCMC):**

MCMC is a class of algorithms used for sampling from complex probability distributions, which is often encountered in Bayesian statistical inference. The primary goal is to generate samples from the posterior distribution of model parameters given observed data. The central idea is to construct a Markov Chain that, when run for a sufficiently long time, converges to the target posterior distribution.

#### **Markov Chain Values and Their Role:**

-   Markov Property:

    -   A Markov Chain is a sequence of random variables where the probability distribution of each variable depends only on the preceding variable. This is known as the Markov property.

    -   In the context of MCMC, Markov Chains are constructed to explore the parameter space of a Bayesian model. Each value in the chain represents a set of parameter values.

-   **Sampling from the Posterior:**

    -   MCMC methods generate a sequence of parameter values from the posterior distribution. Each value in the Markov Chain is a sample from the posterior, and the chain is constructed in a way that it explores the high-probability regions of the parameter space.

-   **Convergence and Mixing:**

    -   The quality of MCMC sampling depends on the convergence and mixing properties of the Markov Chain.

    -   Convergence ensures that the chain has reached a stationary distribution, and further samples provide information about the target posterior.

    -   Mixing refers to how effectively the chain explores the parameter space. A well-mixing chain moves easily between different regions of high posterior probability.

-   **Trace Plots and Diagnostics**:

    -   Practitioners often examine trace plots, which show the values of parameters in the Markov Chain over iterations. These plots help assess convergence and identify potential issues.

    -   Diagnostic tools, such as the Gelman-Rubin statistic, effective sample size, and autocorrelation plots, are used to evaluate the quality of the Markov Chain values.

-   **Inference and Uncertainty Quantification**:

    -   Once a well-behaved Markov Chain is obtained, the values in the chain can be used for Bayesian inference.

    -   Posterior summaries, such as mean, median, and credible intervals, can be computed from the Markov Chain values to quantify uncertainty in parameter estimates.

-   **Bayesian Updating**:

    -   In a Bayesian framework, new data can be incorporated by updating the Markov Chain using the posterior distribution based on the observed data. This allows for iterative learning and model refinement.

#### **Example:**

Consider a Bayesian linear regression model with unknown parameters 0 and 1. MCMC methods could generate a Markov Chain of values for  0 and 1based on observed data. The chain's convergence and mixing properties, as well as the distribution of values in the chain, provide insights into the uncertainty associated with the parameter estimates.

In summary, Markov Chain values are the backbone of MCMC algorithms, enabling Bayesian practitioners to sample from complex posterior distributions, conduct inference, and quantify uncertainty in model parameters.

**The RStan Package**

RStan is an R interface to Stan, which is a probabilistic programming language for Bayesian statistical modeling and high-performance statistical computation. RStan allows users to specify Bayesian models using the Stan language and then perform Bayesian inference using Markov Chain Monte Carlo (MCMC) methods. Keep in mind that software packages may have undergone updates or changes since then, so it's a good idea to check for the latest information.

Here's a general overview of the RStan package and its main functionalities:

-   **Installation**:

    -   To use RStan, you need to install both RStan and the Stan software. Instructions for installation can be found on the official RStan website.

-   **Model Specification**:

    -   RStan allows users to specify Bayesian models using the Stan modeling language. Stan provides a flexible and expressive language for defining hierarchical models, specifying priors, likelihoods, and performing Bayesian inference.

-   **Compilation**:

    -   Once the Stan model is defined in R, it needs to be compiled. RStan facilitates the compilation process, translating the Stan model code into C++ code for efficient computation.

-   **Sampling and Inference**:

    -   RStan uses MCMC algorithms, particularly the No-U-Turn Sampler (NUTS), for sampling from the posterior distribution of model parameters. Users can control the sampling process, such as the number of iterations and chains, through RStan functions.

```{=html}
<!-- -->
```
-   **Posterior Analysis**:

    -   After sampling, RStan provides tools for posterior analysis. Users can examine trace plots, summary statistics, and other diagnostics to assess the convergence and performance of the Markov Chain.

-   Visualization:

    -   RStan includes functions for visualizing results, such as posterior density plots, trace plots, and other diagnostic plots to aid in model assessment.

-   Integration with R Ecosystem:

    -   RStan seamlessly integrates with the broader R ecosystem, allowing users to leverage R's extensive capabilities for data manipulation, visualization, and reporting.

-   Parallelization:

    -   RStan supports parallelization, allowing users to run multiple chains in parallel for faster computation.

Here's a simple example of how you might use RStan to fit a Bayesian linear regression model:

```{r}
# Install and load the RStan package
# install.packages("rstan")
# library(rstan)
# Define the Stan model
# stan_code <- "
# data {
#  int<lower=0> N; # Number of observations
#  vector[N] x; # Predictor variable
#  vector[N] y; # Response variable
# }
# parameters {
#  real alpha; # Intercept
#  real beta; # Slope
#  real<lower=0> sigma; # Standard deviation of the errors
# }
# model {
#  y ~ normal(alpha + beta * x, sigma); # Likelihood
# }
#  Additional blocks for priors and other specifications can be added
# Compile the Stan model
# stan_model <- stan_model(model_code = stan_code)
# Create data list
# data_list <- list(N = length(data$y), x = data$x, y = data$y)
# Run MCMC sampling
# stan_fit <- sampling(stan_model, data = data_list, iter = 2000, chains = 4)

# Print summary
# print(stan_fit)

```
